---
title: Research on NFL Play Data
author: Tommy O'Neil
description: I am aiming to look at a correlation between the distance that a team has to go in order to get a first down and the yards that they end up gaining. 
output:
  distill::distill_article:
    self_contained: false
---

The play result is the outcome variable, yards to go is explanatory

## Project thoughts

I am interested in exploring data related to 

```{r}
library(tidyverse)
plays_data <- read.csv("plays.csv")

plays_data <- plays_data |>
  mutate(plays_data$yardsToGo)
plays_data

plot <- plays_data |>
  filter(yardsToGo == "Remaining Yards") |>
  ggplot(mapping = aes(x = plays_data$yardsToGo)) + geom_histogram()
hist(x = plays_data$yardsToGo, main = "Yards to Go")
```

The above graph is an example of the average number of yards to gain when a team is facing a third down play, which means they have one opportunity to convert or they will likely have to punt, or in other words give the ball away. 

```{r}
plays_data <- plays_data |>
  mutate(plays_data$playResult)
plays_data

plot <- plays_data |>
  filter(yardsToGo == "Remaining Yards") |>
  ggplot(mapping = aes(x = plays_data$playResult)) + geom_histogram()
hist(x = plays_data$playResult, main = "Play Result")
```
histogram of distribution of outcome variable

Subsequently, this histogram shows the average result of these plays, where you can see how the teams usually fare when they are faced with the tall task of trying to convert a third down play. 

```{r}

library(ggplot2)


ggplot(plays_data, aes(x = yardsToGo, y = playResult)) +
  geom_point() +
  xlab("Yards to Go") +
  ylab("Result of the Play")

```
Bivariate plot with explanatory variable on x-axis and outcome variable on y-axis
```{r}
library(broom)

model <- lm(plays_data$playResult ~ plays_data$yardsToGo, plays_data = plays_data)

results <- tidy(model)

results
```
regression results for multivariate model
```{r}

```

