---
title: Research on NFL Plays Data
author: Tommy O'Neil
description: The purpose of this research is to look at a potential correlation between the distance that a team has to go in order to get a first down and the yards that they end up gaining. I always found it interesting while watching football on how the perceived goal that they are trying to achieve ultimately effects the actual yardage they accumulate. It seems that there is sometimes a pronounced effect on strategy and resulting yardage gained when it comes to yards needed to obtain the first down. 
output:
  distill::distill_article:
    self_contained: false
---

Introduction

Is there any correlation between distance to the first down marker in football and the resulting yardage gained? In this study I plan to take a set of data that has a vast amount of plays recorded over the past two decades from numerous games, all of which have the yardage that a specific team had to gain to get a first down, and the resulting yardage gained. All NFL teams have different strategies to an extent, but as an avid football watcher, I have always found it interesting to look at the specific distance that a team has to go in order to get a first down, and then see the result of the play. I also plan to tie in the specific down that is being played through a multivariate analysis to see if the down has even more of an effect on the result of the play. I hypothesize that there will ultimately be no statistical significance to the correlation I find between my variables, as it will show the unpredictability of the game of football. 

Data and Research Design

For my data, I was able to obtain a data set that contained extensive data on plays, one of the more interesting facets of the data being that they are all passing plays. What makes this interesting is that run plays are usually less strategic plays, used to gain a couple of yards here or there, while pass plays are consistently used to try and create big plays or to make a larger impact on the game. Pass plays are also typically used when there is a larger distance that the offense needs to gain in order to get a first down. For my research design, I will be creating multiple plots of the data, and then eventually doing a regression to see if I can discern any correlation from the outcome variable which will be the 'play result' versus the explanatory variable which will be the 'yards to go'. The research design being used would be differences in differences. The variables are presented through a statistical model of plays gathered over the course of multiple NFL seasons. 



```{r, message = FALSE, echo = FALSE}
library(tidyverse)
plays_data <- read.csv("plays.csv")

```


```{r, message = FALSE, echo = FALSE}
plays_data <- plays_data |>
  mutate(plays_data$yardsToGo)
```

```{r}



plot <- plays_data |>
  filter(yardsToGo == "Remaining Yards") |>
  ggplot(mapping = aes(x = plays_data$yardsToGo)) + geom_histogram()
hist(x = plays_data$yardsToGo, main = "First Down Yardage", xlab = "Yards to Go", ylab = "Frequency")
```



The above graph is an example of the average number of yards to gain per each play in the data set, as is displayed through this graph, the most common yardage is typically around 10 yards as this is what each first down play in football begins with.


```{r, message = FALSE, echo = FALSE}
plays_data <- plays_data |>
  mutate(plays_data$playResult)
```

```{r}


plot <- plays_data |>
  filter(yardsToGo == "Remaining Yards") |>
  ggplot(mapping = aes(x = plays_data$playResult)) + geom_histogram()
hist(x = plays_data$playResult, main = "Results of Plays", xlab = "Play Result in Yards", ylab = "Frequency of Result")
  
```


Subsequently, this histogram shows the average result of these plays, where you can see how the graphs differ greatly when it comes to yards gained on the play. The interesting part of this graph is how you can go into the negatives which there seems to actually be a level of frequency to.  

```{r}

library(ggplot2)


ggplot(plays_data, aes(x = yardsToGo, y = playResult)) +
  geom_point() +
  xlab("Yards to Go") +
  ylab("Result of the Play")

```



Looking at this plot, it is the combination of the two graphs above, where I combined both the playy result, along with the yards to go in terms of that specific play. As is typical within a football game, there are some obvious outliers when it comes to this particular graph, however it is interesting to see that there seems to be more of a consistency when it comes to yards to go versus the result of the play. This is just an example of how unpredictable the game of football can be when it comes down to it, which will ultimately come into play when discussing the results of these tests. 


```{r}
library(broom)

model <- lm(plays_data$playResult ~ plays_data$yardsToGo, plays_data = plays_data)

results <- tidy(model)

results
```


The interesting part of this tibble is the difference in the estimate between the intercept versus the yards to go, where you have a very high value for the intercept compared to a very low value for the yards to go. 



```{r}


model_formula <- plays_data$playResult ~ plays_data$yardsToGo + plays_data$down 

fit <- lm(model_formula, plays_data)


predictions <- predict(fit, plays_data = plays_data)


summary(fit)

ggplot(plays_data, aes(x = yardsToGo, y = playResult)) +
  geom_point() +
  geom_line(aes(y = predictions), color = "blue") +
  xlab("Yards to Go") +
  ylab("Result of the Play")
```



This is the multivariate regression model of the yards to go, result of the play, and the down to go in question. The regression serves to show if there is any type of correlation between the variables. Because of the vast amount of data points that are being used it is difficult to tell whether or not the correlation is statistically significant or not simply by looking at it, but this is something that will be discussed within the results part of this study. 

Results

When looking at the results of the study, it is first important to look at the limitations posed by the data set. It is an exceptionally large set of plays which in most cases would be a good thing for a data set, but it makes it more difficult to make an assumption on correlation by simply observing. By looking at the p-values provided by the regression analysis which are 7.439589e-188 and 4.175357e-08 it is difficult to surmise correlation as the values are extremely small. This ultimately means that it would be likely that the null hypothesis would be rejected by using this data set. The statistical significance of the estimated coefficient which are 0.06091 for yards to go and -0.74883 for down to go, means that when the yards to go increases, this usually means the play result yardage increases, whereas when the down to go increases, this results in the play yardage going down. In theory, as someone who has prior football knowledge, this does make logical sense as teams who need more yards will likely attempt to gain more yards, and similarly with the increase of the down, they will take less risks if it is unnecessary. By looking at these two coefficients and factoring in that knowledge, I do believe there to be somewhat of a causal relationship between the variables. A potential confounding variable within this data set that should also be noted when correlating the down to go into the data is that these are all passing plays being represented, which are more likely to occur on 2nd and 3rd down rather than 1st down. 



